---
title: "Exploring Unpopular 2022 Spotify Songs"
author: "Harrison Weru"
date: "2025-08-05"
output:
  html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618)
library(tidyverse)
library(gt)
library(gtExtras)
library(patchwork)
```
In this project, I derived insights from the cleaned dataset that I generated earlier.

```{r}
songs <- read.csv("cleaned_unpopular_songs.csv")

knitr::kable(head(songs, n=10))
```

## Which artists have the most unpopular songs? What's the popularity of their songs on average?
I started by selecting each song one time, omitting duplicates.

```{r}
songs_by_artist <- songs %>% 
  select(
    track_artist, 
    track_name,
    popularity
    ) %>% 
  distinct(
    track_artist,
    track_name,
    popularity
    )

knitr::kable(head(songs_by_artist, n=10))
```
Then, I grouped the new dataset by artist to count the number of songs of each artist and then arrange the counts in descending order before taking the top 10. These are the top artists who have most of their songs with popularity of less than ten.

```{r}
unpopular_artists <- songs_by_artist %>%
  group_by(track_artist) %>% 
  count() %>% 
  arrange(desc(n)) %>%
  head(n=10)

knitr::kable(unpopular_artists)
```

The next step was to find the average popularity score of these artists. I used the dataset of unique songs, grouped by artists. I created a new variable that is the average of the popularity score of the songs of an artist then took the unique scores to preserve the artist and the average popularity score only.
```{r}
avg_pop <- songs_by_artist %>% 
  group_by(track_artist) %>% 
  mutate(avg_pop = round(mean(popularity),2)) %>% 
  ungroup %>% 
  select(track_artist, avg_pop) %>%
  arrange(desc(avg_pop)) %>% 
  distinct(track_artist, avg_pop)

knitr::kable(head(avg_pop))
```

Finally, I joined the two datasets such that I have the artist with the most unpopular songs together with their average popularity
```{r}
most_unpopular_artists <- unpopular_artists %>%
  arrange(desc(n)) %>%
  left_join(avg_pop) %>%
  as_tibble()

knitr::kable(most_unpopular_artists)
```

To create the chart, I started with a table of the average popularity scores by artist
```{r}
plot_avg_pop_scores <- most_unpopular_artists %>%
  select(track_artist, 
         avg_pop
         ) %>%
  gt() %>%
  tab_header(title = md("Average Popularity by Artist")) %>%
  cols_label(track_artist="",
             avg_pop=""
             )

plot_avg_pop_scores
```
Then, I created a lolipop chart of the number of songs of each artist
```{r fig.width = 8, out.width="70%", fig.align='center'}
plot_artist_unpop_songs <- ggplot(most_unpopular_artists, aes(x=reorder(track_artist, n), y=n)) +
  geom_segment( aes(x=reorder(track_artist, n), xend=reorder(track_artist, n), y=0, yend=n), size=0.75) +
  geom_point(size=5, color="#1DB954", fill="#1DB954", shape=21, stroke=2) +
  coord_flip() +
  labs(
    title = "Top 10 Artists with the Most Unpopular Songs",
    x = "",
    y = "Number of Songs"
  ) +
  theme(
    plot.margin = grid::unit(c(.5,.5,1.5,.5), "cm"),
    axis.text.y = element_text(
      size = 12, 
      family = "Helvetica",
      hjust = 1
    ),
    axis.text.x = element_text(size = 12, vjust = -1),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = NA),
    plot.title = element_text(
      size = 18,
      hjust = 0.5,
      vjust = 1
    ),
    axis.title.x = element_text(
      hjust = 0.5, 
      size=14, 
      vjust=-2
    ),
    axis.title.y = element_text(
      hjust = 0.5, 
      size=14, 
      vjust = 4
    ),
    text = element_text(
      family = "Helvetica", 
      face = "bold"
    )
  )

plot_artist_unpop_songs

```

Finally, I combined the two in the same plot to have a complete plot.
```{r fig.width = 12, out.width="100%", fig.align='center'}
plot_artist_unpop <- plot_artist_unpop_songs +
  inset_element(plot_avg_pop_scores, 
                left = 0.7, 
                bottom = 0.5, 
                right = 0.75, 
                top = 0.6)

plot_artist_unpop
```

## How does popularity score varies across music keys? Can danceability explain the variation?

```{r fig.width = 8, out.width="70%", fig.align='center'}
popularity_key <- ggplot(songs, 
       aes(
         x = popularity,
         y = key_fct
         )
       ) +
  geom_jitter(
    aes(color = danceability)
    ) +
  geom_boxplot(
    aes(group = key_fct), 
    color = "black", 
    alpha = 0.75, 
    width = 0.5, 
    outlier.shape = NA
  ) +
  scale_color_gradientn(
    colours = c(
      "black", 
      "blue", 
      "magenta", 
      "yellow"
    ),
    name = "Danceability"
  ) +
  labs(
    title = "Track Popularity for Each Music Key",
    x = "Track Popularity",
    y = "Music Key"
  ) +
  theme(
    legend.text = element_text(colour = "grey20"),
    plot.margin = grid::unit(c(.5,.5,1.5,.5), "cm"),
    axis.text.y = element_text(
      size = 10, 
      family = "Helvetica",
      hjust = 1
    ),
    axis.text.x = element_text(size = 10, vjust = -1),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = NA),
    plot.title = element_text(
      hjust = 0.5,
      vjust = 1
    ),
    axis.title.x = element_text(
      hjust = 0.5, 
      size=12, 
      vjust=-2
    ),
    axis.title.y = element_text(
      hjust = 0.5, 
      size=12, 
      vjust = 4
    ),
    text = element_text(
      family = "Helvetica", 
      face = "bold"
    )
  )

popularity_key
```