---
title: "Cleaning OICA Perssengers Cars Sales 2019-2024 Data"
author: "Harrison Weru"
date: "2025-07-11"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(readxl)
library(janitor)
```

## Loading the dataset
```{r}
pc_sales <- read_xlsx("pc_sales_2024.xlsx", skip = 3)
```

## viewing the dataframe to note initial problems
```{r}
str(pc_sales)
dim(pc_sales)
```

## Notable issues:
1. The column names should be cleaned 
2. The dataset is in wide format instead of long format
3. The first column contains both regions and countries.

### 1. Renaming the columns and removing NAs
```{r}
pc_sales <- pc_sales %>% 
  select(1:7) %>%
  rename(
    "country" = "REGIONS/COUNTRIES",
    "2019" = "Q1-Q4 2019",
    "2020" = "Q1-Q4 2020",
    "2021" = "Q1-Q4 2021",
    "2022" = "Q1-Q4 2022",
    "2023" = "Q1-Q4 2023",
    "2024" = "Q1-Q4 2024"
  ) %>%
  remove_empty()

head(pc_sales, n=20)
```

### 2. Converting the dataframe to wide format
```{r}
pc_sales <- pc_sales %>%
  pivot_longer(cols = 2:7, names_to = "year", values_to = "sales") %>%
  mutate(year = as.numeric(year))
head(pc_sales, n=20)
```


### 3. Creating a separate dataframe for country and regions
```{r}
regions <- c (
  "EUROPE",
  "EU 27 countries + EFTA + UK",
  "RUSSIA, TURKEY & OTHER EUROPE",
  "OTHER COUNTRIES/REGIONS",
  "AMERICA",
  "USMCA (former NAFTA)",
  "CENTRAL & SOUTH AMERICA",
  "ASIA/OCEANIA/MIDDLE EAST",
  "ASEAN",
  "AFRICA",
  "ALL COUNTRIES/REGIONS",
  "TOTAL OICA MEMBERS",
  "* including LCV& HCV",
  "OTHER COUNTRIES"
)

pc_sales_region <- pc_sales %>%
  filter(country %in% regions & !is.na(sales)) %>%
  rename("regions" = "country")

head(pc_sales_region, n=20)

pc_sales_country <- pc_sales %>%
  filter(!(country %in% regions)) %>% 
  mutate(country = str_to_title(country)) %>%
  arrange(country, year)

head(pc_sales_country, n=20)
```

## Exporting the cleaned data frames
```{r}
write.csv(pc_sales_country, file = "clean_passenger_car_sales_by_country_2019-2024.csv")

write.csv(pc_sales_region, file = "clean_passenger_car_sales_by_region_2019-2024.csv")
```
